Grove: A Convergent Collaborative Structure-Editor Calculus

### 2021-02-16 grove-paper.tex
replace \multiChild with \conflictHole
tighten up spacing in \conflictHole

\newcommand{\hole}{\ensuremath{\square}} %\textcolor{violet}{\llparenthesis}}\textcolor{violet}{\rrparenthesis}}
\newcommand{\conflictHole}[1]{%
  {\noexpandarg\StrSubstitute{#1}{,}{\textcolor{red}{\;\textbf{|}\;}}[\myargs]
  {\textcolor{red}{\textbf{\{}} \myargs \textcolor{red}{\textbf{\}}}}}}

@@

\newcommand{\Edge}[1]{Edge~#1}
\newcommand{\Vertex}[1]{Vertex~#1}
\newcommand{\figureCode}[1]{\textbf{\texttt{#1}}\vskip1em}
\newcommand{\multiChild}[2]{\textcolor{red}{\{} #1 \textcolor{red}{|} #2 \textcolor{red}{\}}}
\newcommand{\multiVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\curlyveedownarrow_{#1}}}}}
\newcommand{\cycleVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\rcirclearrowleft_{#1}}}}}
\newcommand{\orphanVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\ndownarrow_{#1}}}}}
\newcommand{\parens}[1]{\textcolor{gray}{(}#1\textcolor{gray}{)}}
\newcommand{\otherVertexVskip}{\vskip0.5em}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureMultiChild
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{MultiChildAlice}{R}
\newvertex{MultiChildAlice}{x}
\newedge{MultiChildBob}{R}
\newvertex{MultiChildBob}{y}
\newcommand{\figureMultiChild}{
\begin{figure}
%\hfill
\hskip0.12\columnwidth
%%%%%%%%
\begin{subfigure}[t]{0.21\linewidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (node/NestedParts/cx) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:NestedParts:c}}}};
%%%%
\path (0cm,0cm) node (node/MultiChild/a) [anchor=base] {\strut};
\path [draw,->,alice step] (node/NestedParts/cx) to (-2em, 0cm |- node/MultiChild/a);
\end{tikzpicture}
}%
\label{fig:MultiChild:a}
\figureCode{w * v + x}%
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedPartsBob"'] -> {
    "\vNestedPartsAlice" [> "\eNestedPartsAlice"'],
    "\vDifferentPartsBob" [> "\eDifferentPartsBob"]
   },
   {
    "\vMultiChildAlice" [> "\eMultiChildAlice", >alice edge, alice node]
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\linewidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (node/MultiChild/b) [anchor=base] {\strut};
\path [draw,->,bob step] (node/NestedParts/cx) to[out=15,in=165] (-2em, 0cm |- node/MultiChild/b);
\end{tikzpicture}
}%
\label{fig:MultiChild:b}
\figureCode{w * v + y}%
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedPartsBob"'] -> {
    "\vNestedPartsAlice" [> "\eNestedPartsAlice"'],
    "\vDifferentPartsBob" [> "\eDifferentPartsBob"]
   },
   "\vMultiChildBob" [> "\eMultiChildBob", >bob edge, bob node]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.4\linewidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (node/MultiChild/c) [anchor=base] {\strut};
\path [draw,->,merge step] (node/MultiChild/a) to[out=15,in=165] (-2em, 0cm |- node/MultiChild/c);
\path [draw,->,merge step] (node/MultiChild/b) to (-2em, 0cm |- node/MultiChild/c);
\end{tikzpicture}
}%
\label{fig:MultiChild:c}
\figureCode{w * v + \multiChild{x}{y}}%
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedPartsBob"'anchor=-15] -> {
    "\vNestedPartsAlice" [> "\eNestedPartsAlice"'],
    "\vDifferentPartsBob" [> "\eDifferentPartsBob"]
   },
   "\vMultiChildAlice" [> "\eMultiChildAlice"anchor=153, >merge edge, merge node],
   {},
   "\vMultiChildBob" [> "\eMultiChildBob"anchor=195, >merge edge, merge node]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
%\hfill{}
\caption{Example of Multi-Child Conflicts.}%
\label{fig:MultiChild}
\end{figure}
}

@@

\figureCode{\parens{\hole{} * \hole{}} + \multiChild{\multiVertex{30}}{\multiVertex{32}}%

@@

We merge these edits by including \emph{both} \eMultiChildAlice{} and \eMultiChildBob{}
as edges going out of the \texttt{R} position of \vWrapPlus{}.
This results in \autoref{fig:MultiChild:c}, where there is a multi-child conflict
in the~\texttt{R} position of~\vWrapPlus~(i.e., there are multiple edges for the same position in a vertex).
In the user interface, we flag this conflict with the notation~\conflictHole{x, y}.

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      q    & {}\in Pat & {}::={} & x^u \mid \hole \mid \conflictHole{q, \ldots, q} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      \tau & {}\in Typ & {}::={} & \tau \rightarrow^u \tau \mid Num^u \mid \hole \mid \conflictHole{\tau, \ldots, \tau} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      e    & {}\in Exp & {}::={} & x^u \mid \lambda^u q \mathord{:}\tau.e \mid (e~e)^u \mid n^u \\
           &           &         & ~\mid e~\texttt{+}^u~e \mid e~\texttt{*}^u~e \mid \hole \mid \conflictHole{e, \ldots, e} \mid \multiVertex{u} \mid \cycleVertex{u} \\
    \end{array}
  \]
  \caption{Base syntax, as a grammar, for the lambda calculus that we are considering.}
  \label{fig:Base syntax}
\end{figure}

@@

The language we are modeling is shown in \autoref{fig:Base syntax}, standard for a simply-typed lambda calculus except that all sorts have empty holes (represented by $\hole$) and conflict holes (of the form $\conflictHole{q, \ldots, q}$).
In particular, the argument of a lambda can be either a bound variable or a hole.
Additionally, all non-hole forms have a superscripted identifier $u \in \U$.

@@

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from $v_{38} = (38, \ExpPlus)$ is $\expr'(v_{38}, \PlusLeft) \text{+}^{38} \expr'(v_{38}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1, \ldots, e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $W$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

  \begin{align*}
    \expr'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \conflictHole{\expr(v_1), \ldots, \expr(v_n)} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \expr(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \patt'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \conflictHole{\patt(v_1), \ldots, \patt(v_n)} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \patt(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \type'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \conflictHole{\type(v_1), \ldots, \type(v_n)} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \type(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

@@

We determine $e$ by starting at the root of the graph and including every connected
vertex that has a single parent. In positions where there are no children, we
leave an empty hole $\hole$. In positions where there are multiple children, we
leave a conflict hole of the form $\conflictHole{e_1, \ldots, e_n}$. Whenever we
encounter a vertex with multiple parents, we add its corresponding expression to
the second component $MP$ and leave an indirect reference $\#u$, where $u$ is
the unique identifier of the referenced vertex

>>>>

\newcommand{\hole}{\ensuremath{\square}} %\textcolor{violet}{\llparenthesis}}\textcolor{violet}{\rrparenthesis}}
\newcommand{\conflictHole}[1]{%
{\noexpandarg\StrSubstitute{#1}{,}{\textcolor{red}{\;\textbf{|}\;}}[\myargs]%
{\textcolor{red}{\textbf{\{}}\myargs\textcolor{red}{\textbf{\}}}}}}%

@@

\newcommand{\Edge}[1]{Edge~#1}
\newcommand{\Vertex}[1]{Vertex~#1}
\newcommand{\figureCode}[1]{\textbf{\texttt{#1}}\vskip1em}
\newcommand{\multiVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\curlyveedownarrow_{#1}}}}}
\newcommand{\cycleVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\rcirclearrowleft_{#1}}}}}
\newcommand{\orphanVertex}[1]{\textcolor{red}{\ensuremath{\mathbf{\ndownarrow_{#1}}}}}
\newcommand{\parens}[1]{\textcolor{gray}{(}#1\textcolor{gray}{)}}
\newcommand{\otherVertexVskip}{\vskip0.5em}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \figureMultiChild
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newedge{MultiChildAlice}{R}
\newvertex{MultiChildAlice}{x}
\newedge{MultiChildBob}{R}
\newvertex{MultiChildBob}{y}
\newcommand{\figureMultiChild}{
\begin{figure}
%\hfill
\hskip0.12\columnwidth
%%%%%%%%
\begin{subfigure}[t]{0.21\linewidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (-7.5em,0cm) node (node/NestedParts/cx) [gray,anchor=base west] {{\textbf{\small Fig.~\ref*{fig:NestedParts:c}}}};
%%%%
\path (0cm,0cm) node (node/MultiChild/a) [anchor=base] {\strut};
\path [draw,->,alice step] (node/NestedParts/cx) to (-2em, 0cm |- node/MultiChild/a);
\end{tikzpicture}
}%
\label{fig:MultiChild:a}
\figureCode{w * v + x}%
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedPartsBob"'] -> {
    "\vNestedPartsAlice" [> "\eNestedPartsAlice"'],
    "\vDifferentPartsBob" [> "\eDifferentPartsBob"]
   },
   {
    "\vMultiChildAlice" [> "\eMultiChildAlice", >alice edge, alice node]
   }
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.21\linewidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (node/MultiChild/b) [anchor=base] {\strut};
\path [draw,->,bob step] (node/NestedParts/cx) to[out=15,in=165] (-2em, 0cm |- node/MultiChild/b);
\end{tikzpicture}
}%
\label{fig:MultiChild:b}
\figureCode{w * v + y}%
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedPartsBob"'] -> {
    "\vNestedPartsAlice" [> "\eNestedPartsAlice"'],
    "\vDifferentPartsBob" [> "\eDifferentPartsBob"]
   },
   "\vMultiChildBob" [> "\eMultiChildBob", >bob edge, bob node]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}[t]{0.4\linewidth}
\centering
\caption{
\begin{tikzpicture}[remember picture, overlay]
\path (0cm,0cm) node (node/MultiChild/c) [anchor=base] {\strut};
\path [draw,->,merge step] (node/MultiChild/a) to[out=15,in=165] (-2em, 0cm |- node/MultiChild/c);
\path [draw,->,merge step] (node/MultiChild/b) to (-2em, 0cm |- node/MultiChild/c);
\end{tikzpicture}
}%
\label{fig:MultiChild:c}
\figureCode{w * v + \conflictHole{x,y}}%
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "\vWrapPlus" [> "\eWrapPlus"] -> {
   "\vSimpleTimes" [> "\eNestedPartsBob"'anchor=-15] -> {
    "\vNestedPartsAlice" [> "\eNestedPartsAlice"'],
    "\vDifferentPartsBob" [> "\eDifferentPartsBob"]
   },
   "\vMultiChildAlice" [> "\eMultiChildAlice"anchor=153, >merge edge, merge node],
   {},
   "\vMultiChildBob" [> "\eMultiChildBob"anchor=195, >merge edge, merge node]
  }
 }
};
\end{tikzpicture}
\end{subfigure}
%%%%%%%%
%\hfill{}
\caption{Example of Multi-Child Conflicts.}%
\label{fig:MultiChild}
\end{figure}
}

@@

\figureCode{\parens{\hole{} * \hole{}} + \conflictHole{\multiVertex{30},\multiVertex{32}}%

@@

We merge these edits by including \emph{both} \eMultiChildAlice{} and \eMultiChildBob{}
as edges going out of the \texttt{R} position of \vWrapPlus{}.
This results in \autoref{fig:MultiChild:c}, where there is a multi-child conflict
in the~\texttt{R} position of~\vWrapPlus~(i.e., there are multiple edges for the same position in a vertex).
In the user interface, we flag this conflict with the notation~\conflictHole{x,y}.

@@

\begin{figure}
  \[
    \arraycolsep=0pt
    \begin{array}{lrlll}
      q    & {}\in Pat & {}::={} & x^u \mid \hole \mid \conflictHole{q,\ldots,q} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      \tau & {}\in Typ & {}::={} & \tau \rightarrow^u \tau \mid Num^u \mid \hole \mid \conflictHole{\tau,\ldots,\tau} \mid \multiVertex{u} \mid \cycleVertex{u} \\
      e    & {}\in Exp & {}::={} & x^u \mid \lambda^u q \mathord{:}\tau.e \mid (e~e)^u \mid n^u \\
           &           &         & ~\mid e~\texttt{+}^u~e \mid e~\texttt{*}^u~e \mid \hole \mid \conflictHole{e,\ldots,e} \mid \multiVertex{u} \mid \cycleVertex{u} \\
    \end{array}
  \]
  \caption{Base syntax, as a grammar, for the lambda calculus that we are considering.}
  \label{fig:Base syntax}
\end{figure}

@@

The language we are modeling is shown in \autoref{fig:Base syntax}, standard for a simply-typed lambda calculus except that all sorts have empty holes (represented by $\hole$) and conflict holes (of the form $\conflictHole{q,\ldots,q}$).
In particular, the argument of a lambda can be either a bound variable or a hole.
Additionally, all non-hole forms have a superscripted identifier $u \in \U$.

@@

Let us now examine each of the four components in more detail.
The first component $e_r$ is the expression derived from the root vertex $v_0$.
\autoref{fig:Decomposition expr} defines $\expr(v)$ wich determines the expression derived from $v$.
There is one rule for each constructor form.
For example, the expression derived from $v_{38} = (38, \ExpPlus)$ is $\expr'(v_{38}, \PlusLeft) \text{+}^{38} \expr'(v_{38}, \PlusRight)$.
In each child position, $\expr'(v, p)$ determines the sub-expression derived from the children of $v$ at position $p$.
An empty hole $\hole$ appears if there are no children at the given position.
A conflict hole $\conflictHole{e_1,\ldots,e_n}$ appears if there are multiple children.
If there is one child, the expression is determined recursively unless that child appears in $MP$ or $W$, discussed below, in which case we leave a reference, $\multiVertex{u}$ or $\cycleVertex{u}$ respectively.
For the $\ExpLam$ case, we define $\patt(q)$ and $\type(\tau)$ similarly.

@@

  \begin{align*}
    \expr'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \conflictHole{\expr(v_1),\ldots,\expr(v_n)} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \expr(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \patt'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \conflictHole{\patt(v_1),\ldots,\patt(v_n)} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \patt(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

  \begin{align*}
    \type'(v,p) = \begin{cases}
      \hole & \children(v,p) = \varnothing \\
      \conflictHole{\type(v_1),\ldots,\type(v_n)} & \children(v,p) = \{v_1, \ldots, v_n\} \\
      \multiVertex{u} & \children(v,p) = \{v'\} \land |\parents(v')| > 1 \\
      \cycleVertex{u} & \children(v,p) = \{v'=(u,k)\} \land |\parents(v')| = 1 \\
          & \phantom{\children(v,p)} \land v' = \min(\ancestors(v')) \\
      \type(v') & \text{otherwise} \\
    \end{cases}
  \end{align*}

@@

We determine $e$ by starting at the root of the graph and including every connected
vertex that has a single parent. In positions where there are no children, we
leave an empty hole $\hole$. In positions where there are multiple children, we
leave a conflict hole of the form $\conflictHole{e_1,\ldots,e_n}$. Whenever we
encounter a vertex with multiple parents, we add its corresponding expression to
the second component $MP$ and leave an indirect reference $\#u$, where $u$ is
the unique identifier of the referenced vertex
